smooth-sea

### 2018-05-29 index.rmd
presentation version

#create neighbor file that INLA takes for hyperparameter args
nb2INLA("./output/seatract.graph", sea_adj)
sea_df$idqtr <- factor(sea_df$listingQtr, ordered = T)
sea_df$idqtr1 <- sea_df$idqtr
sea_df$idtractqtr <- paste(sea_df$idtract, sea_df$idqtr)
sea_df$idtract1 <- sea_df$idtract
sea_df$Qtr1 <- sea_df$Qtr

@@

summary(m.int)
sea_df$int_Med <- m.int$summary.linear.predictor[, "0.5quant"]
sea_df$int_SD <- m.int$summary.linear.predictor[, "sd"]
sea_df$int_postWidth <- m.int$summary.linear.predictor[, "0.975quant"] - m.int$summary.linear.predictor[,"0.025quant"]
sea_df$int_Eff <- m.int$summary.random$idtract[sea_df$idtract, "0.5quant"]

@@

summary(m.nsar1) #NB: does not improve fit to use random effect time specification, using fixed effects
sea_df$nsar1_Med <- m.nsar1$summary.linear.predictor[, "0.5quant"]
sea_df$nsar1_SD <- m.nsar1$summary.linear.predictor[, "sd"]
sea_df$nsar1_postWidth <- m.nsar1$summary.linear.predictor[, "0.975quant"] - m.nsar1$summary.linear.predictor[,"0.025quant"]
sea_df$nsar1_Eff <- m.nsar1$summary.random$idqtr[sea_df$idqtr, "0.5quant"]

#spatial random effect model (BYM 1991) with seasonal dummies
form.bym <- log(med1B) ~ 1 + Qtr +
  f(idqtr, model = "iid") +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "./output/seatract.graph")

@@

#spatio-temporal smoothing model, AR(1) process with linear space-time interaction (Bernadelli 1995)
form.spt <- log(med1B) ~ 1 +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "./output/seatract.graph") +
  f(idqtr, model = "ar", order = 1) +
  f(idtractqtr, model = "iid")

@@

The goal of this research is to a.) understand the temporal dynamics of rent estimates in Seattle and b.) forecast the current quarter's rent levels based off of the prior periods. The focal series of models regress median one-bedroom rent asked values on different specifications of the panel's correlation structure (i.e. temporal and spatial). All of the candidate models are estimated with [integrated nested Laplace approximations](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) (INLA) using the default, weakly-informative priors for all model hyperparameters. Throughout the following analyses, the training data are `r min(sea_df$listingQtr)` up to the prior quarter (i.e. `r as.yearqtr(Sys.Date())-.25`). The test period is a forecast for the current period and includes comparison to the appropriate median rent estimates for data observed so far in this period.

@@

ns_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = ns_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "C. Non-Spatial RE") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

nsar1_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = nsar1_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "D. Non-Spatial AR(1) RE") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

bym_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = bym_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "E. Spatial RE") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

@@

### Spatial RE

@@

### Spatiotemporal AR(1) RE

@@

Model abbr.                   | Description
----------------------------- | -----------------------------------------------
**int**    | Quarter fixed intercept
           | `r format(form.int, )`
**ns**     | Non-spatial tract random effect for each tract, quarter fixed intercept
           | `r format(form.ns)`
**nsar1**  | Non-spatial tract random effect model, AR(1) random effect for quarter, i.i.d. random effect for quarter 
           | `r format(form.nsar1)`
**bym**    | Spatial intrinsic conditional autoregressive (ICAR) tract random-effect with quarter fixed intercept 
           |  `r format(form.bym)`
**spt**    | Spatial intrinsic conditional autoregressive (ICAR) tract random-effect, AR(1) random effect for quarter, i.i.d. random effect for quarter and tract-quarter
           | `r format(form.spt)`

@@

@@

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = nsar1_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()

@@

### Spatial RE

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = bym_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()

@@

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = spt_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()

>>>>

#create neighbor file that INLA takes for hyperparameter args
nb2INLA("./output/seatract.graph", sea_adj)
sea_df$idqtr <- factor(sea_df$listingQtr, ordered = T)
sea_df$idqtr1 <- sea_df$idqtr
sea_df$idtractqtr <- paste(sea_df$idtract, sea_df$Qtr)
sea_df$idtract1 <- sea_df$idtract
sea_df$Qtr1 <- sea_df$Qtr

@@

summary(m.int)
sea_df$int_Med <- m.int$summary.linear.predictor[, "0.5quant"]
sea_df$int_SD <- m.int$summary.linear.predictor[, "sd"]
sea_df$int_postWidth <- m.int$summary.linear.predictor[, "0.975quant"] - m.int$summary.linear.predictor[,"0.025quant"]

@@

summary(m.nsar1) #NB: does not improve fit to use random effect time specification, using fixed effects
sea_df$nsar1_Med <- m.nsar1$summary.linear.predictor[, "0.5quant"]
sea_df$nsar1_SD <- m.nsar1$summary.linear.predictor[, "sd"]
sea_df$nsar1_postWidth <- m.nsar1$summary.linear.predictor[, "0.975quant"] - m.nsar1$summary.linear.predictor[,"0.025quant"]
sea_df$nsar1_Eff <- m.nsar1$summary.random$idtract[sea_df$idtract, "0.5quant"]

#spatial random effect model (BYM 1991) with AR(1)
form.bym <- log(med1B) ~ 1 +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "./output/seatract.graph") +
  f(idqtr, model = "ar1") +
  f(idqtr1, model = "iid")

@@

#spatio-temporal smoothing model, with AR(1) and linear space-time interaction (Bernadelli 1995)
form.spt <- log(med1B) ~ 1 +
  f(idtract, model = "bym", #ICAR spatial RE for tract neighbors + IID RE for tract
    scale.model = T,
    graph = "./output/seatract.graph") +
  f(idqtr, model = "ar1") +
  f(idqtr1, model = "iid") +
  f(idtractqtr, model = "iid")

@@

The goal of this research is to a.) understand the temporal dynamics of rent estimates in Seattle and b.) forecast the current quarter's rent levels based off of the prior periods. The focal series of models regress median one-bedroom rent asked values on different specifications of the panel's correlation structure (i.e. temporal and spatial). All of the candidate models' posterior distributions are estimated with [integrated nested Laplace approximations](https://folk.ntnu.no/hrue/r-inla.org/papers/inla-rss.pdf) (INLA) using the default, weakly-informative priors for all model hyperparameters. Throughout the following analyses, the training data are `r min(sea_df$listingQtr)` up to the prior quarter (i.e. `r as.yearqtr(Sys.Date())-.25`). The test period is a forecast for the current period and includes comparison to the appropriate median rent estimates for data observed so far in this period.

@@

ns_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = ns_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "C. Non-Spatial") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

nsar1_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = nsar1_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "D. Non-Spatial AR(1)") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

bym_df <- sea_df %>%
  group_by(listingQtr) %>%
  mutate(value = bym_Med,
         qtrDev = value-mean(value, na.rm=T),
         model = "E. Spatial AR(1)") %>%
  dplyr::select(listingQtr, value, GISJOIN, qtrDev, model)

@@

### Spatial AR(1)

@@

@@

Model abbr.                   | Description
----------------------------- | -----------------------------------------------
**int**    | Quarter fixed intercept
           | `r format(form.int, )`
**ns**     | Non-spatial tract random effect for each tract, quarter fixed intercept
           | `r format(form.ns)`
**nsar1**  | Non-spatial tract random effect, AR(1) random effect for prior quarter, i.i.d random effect for current quarter
           | `r format(form.nsar1)`
**bym**    | Spatial intrinsic conditional autoregressive (ICAR) tract random-effect, non-spatial i.i.d tract random effect, AR(1) random effect for prior quarter, i.i.d random effect for current quarter
           |  `r format(form.bym)`
**spt**    | Spatial intrinsic conditional autoregressive (ICAR) tract random-effect, non-spatial i.i.d tract random effect, AR(1) random effect for prior quarter, i.i.d random effect for current quarter, i.i.d. random effect for tract-quarter (space-time interaction)
           | `r format(form.spt)`

@@

### Hyperparameters

```{r, echo=F}
knitr::kable(summary(m.nsar1)$hyperpar, caption = "Non-Spatial AR(1)")
```

```{r, echo=F}
knitr::kable(summary(m.bym)$hyperpar, caption = "Spatial AR(1)")
```

```{r, echo=F}
knitr::kable(summary(m.spt)$hyperpar, caption = "Spatiotemporal AR(1)")
```

@@

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = nsar1_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(limits = c(800, 2600), labels = scales::dollar) +
  scale_y_continuous(limits = c(800, 2600), labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()

@@

### Spatial AR(1)

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = bym_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(limits = c(800, 2600), labels = scales::dollar) +
  scale_y_continuous(limits = c(800, 2600), labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()

@@

```{r, echo=F}
avp_scatter <- ggplot(sea_df, aes(x = actualRent, y = spt_Med, size = n1B, frame = listingQtr)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(limits = c(800, 2600), labels = scales::dollar) +
  scale_y_continuous(limits = c(800, 2600), labels = scales::dollar) +
  xlab("Actual") +
  ylab("Predicted") +
  scale_size_area() +
  coord_equal() +
  theme_minimal()
